---
title: "Peer Assessment 1 - Analyzing activity monitoring data"
author: "Martynas Asipauskas"
date: "14 June 2015"
output: 
    html_document:
        keep_md: true
---
```{r, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path = "figures/")
```

## Abstract
This is a very high level analysis of activity monitoring data provided in `activity.zip` dataset.
Analysis is being done as a peer assessed coursework for Coursera - Reproducible Research course. 

`Activity` dataset contains a set of measurements collected between October and November, 2012 from Anonymous individual and contains number of steps taken in 5 minute intervals every day.

Summary of the dataset can be seen bellow:
```{r}
library(data.table)

activity <- read.csv(unz("activity.zip", filename = "activity.csv"))
activity$interval <- as.factor(activity$interval)
activity$date <- as.Date(activity$date)
activity <- data.table(activity)
summary(activity)
```

## What is mean total number of steps taken per day?
You can see the histogram of total number of steps taken per day between October and November bellow. 
```{r}
numberOfStepsPerDay <- as.data.frame(activity[, sum(steps, na.rm = TRUE), by = date])
hist(numberOfStepsPerDay$V1, main = "Steps per day", xlab = "Steps per day", ylab = "Days", col = "#003366", border = "red")
```

```{r}
meanNumberOfStepsTaken <- round(mean(numberOfStepsPerDay$V1), 2)
meanNumberOfStepsTaken

medianNumberOfStepsTaken <- round(median(numberOfStepsPerDay$V1), 2)
medianNumberOfStepsTaken
```

The mean of steps taken in any given day is: *`r meanNumberOfStepsTaken`* steps. The meadian is: *`r medianNumberOfStepsTaken`* steps.

## What is the average daily activity pattern?
```{r}
averageNumberOfStepsPerInterval <- as.data.frame(activity[, mean(steps, na.rm = TRUE), by = interval])
plot(averageNumberOfStepsPerInterval$interval, averageNumberOfStepsPerInterval$V1, type = "l", xlab = "5-min interval", 
    ylab = "Number of Steps", main = "Average number of steps taken per 5-min interval", 
    col = "#003366", border = "red")
lines(averageNumberOfStepsPerInterval$interval, averageNumberOfStepsPerInterval$V1, col = "#003366")
```

```{r}
maxInterval <- averageNumberOfStepsPerInterval[which.max(averageNumberOfStepsPerInterval$V1),]$interval
```

The 5-min interval, during which the most steps are taken is: *`r maxInterval`*.

# Imputing missing values
The `Activity` dataset contains quite a few missing values.

```{r}
nas <- sum(is.na(activity$steps))
nas
```

As can be seen from the summary above, the total number of missing values for steps is: *`r nas`*. 

To fill in the missing values in this dataset, we will use a very simple strategy, where `NA` value will be replaced by the mean of the value in that interval.

```{r}
library(plyr)
activityNoNa <- data.table(ddply(activity, .(interval), function(df) {df$steps[is.na(df$steps)] <- mean(df$steps, na.rm=TRUE); return(df)}))
```

```{r}
numberOfStepsPerDayNoNa <- as.data.frame(activityNoNa[, sum(steps, na.rm = TRUE), by = date])
hist(numberOfStepsPerDayNoNa$V1, main = "Steps per day", xlab = "Steps per day", ylab = "Days", col = "#003366", border = "red")
```

```{r}
meanNumberOfStepsTakenNoNa <- round(mean(numberOfStepsPerDayNoNa$V1), 2)
meanNumberOfStepsTakenNoNa

medianNumberOfStepsTakenNoNa <- round(median(numberOfStepsPerDayNoNa$V1), 2)
medianNumberOfStepsTakenNoNa
```

Once `NA` values have been filled in, the resulting mean of steps taken in any given day is: *`r meanNumberOfStepsTakenNoNa`* steps. The meadian is: *`r medianNumberOfStepsTakenNoNa`* steps. Which marks an increase of the mean by: *`r (meanNumberOfStepsTakenNoNa - meanNumberOfStepsTaken)`* and median by: *`r (medianNumberOfStepsTakenNoNa - medianNumberOfStepsTaken)`* steps.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
library(lattice)
activityNoNa$weekend <- as.factor(weekdays(activityNoNa$date) %in% c("Saturday", "Sunday"))
activityNoNa$weekend <- mapvalues(activityNoNa$weekend, from = c(TRUE, FALSE), to = c("Weekend", "Weekday"))

stepsPerDay <- aggregate(steps ~ interval + weekend, data = activityNoNa, mean)
xyplot(steps ~ interval | weekend, stepsPerDay, type = "l", layout = c(1, 2), 
    xlab = "Interval", ylab = "Number of steps")
```
