---
title: "Statistical Inference Course Project - Part 2"
author: "Martynas Asipauskas"
date: "21 June 2015"
output: pdf_document
---
## Overview
As a second part in statistical inference course project. In this report we will look into exploring `ToothGrowth` dataset and performing basic inferential analysis on it.
Specifically we wil be trying to look into the length of teeth in each of guinea pigs in response to three different doses of Vitamin C (0.5 mg, 1 mg, 2 mg) and two delivery methods: orange juicce and ascorbic acid.

## Exploring the data set
Lets load the data set first and have a quick look at the contents of the dataset
```{r}
library(ggplot2)
library(datasets)
data(ToothGrowth)

data <- ToothGrowth
data$dose <- as.factor(data$dose)

summary(data)
ggplot(data, aes(x = dose, y = len)) + geom_boxplot(aes(fill = dose)) + facet_grid(.~ supp)
```

## Comparing tooth growth by `supp` and `dose`
We will compare the tooth growth by testing the following hypothesis:

- If different type of suplement has any impact to the teeth length
- If having different doses of Vitamin C does have any impact to resulting teeth length.

In both cases we are going to be trying to reject the null hypothesis, that type of supplement and dose of Vitamin C has no impact on resulting teeth length in guinea pigs.    

### Different type of suplement impact to the teeth length
```{r}
t.test(len ~ supp, data = data)
```
As p-value of the test is 0.06 and resulting confidence interval contains zero, 
we can't reject our null hypothesis that differnt supplement methods have no effect on resulting tooth length in guinea pigs.

### Different doses of Vitamin C impact to the teeth length
We are going to start by spliting our data set into 3 sections to test impact of dose increase from 0.5 mg to 1 mg, from 0.5 mg to 2 mg and 1 mg to 2 mg.
```{r}
increase_05_to_1mg <- data[data$dose %in% c(0.5, 1.0), ]
increase_05_to_2mg <- data[data$dose %in% c(0.5, 2.0), ]
increase_1_to_2mg <- data[data$dose %in% c(1.0, 2.0), ]
```

Increase of Vitamin C dose from 0.5 mg to 1 mg.
```{r}
t.test(len ~ dose, data = increase_05_to_1mg)
```

Increase of Vitamin C dose from 0.5 mg to 2 mg.
```{r}
t.test(len ~ dose, data = increase_05_to_2mg)
```

Increase of Vitamin C dose from 1 mg to 2 mg.
```{r}
t.test(len ~ dose, data = increase_1_to_2mg)
```

As for all 3 tests confidence interval does not include 0 and p-value's are less than 0.05 (5%) we can infer that increase in Vitamin C does, does lead to increased teeth length in guinea pigs. Therefore we reject our null hypothesis (that there is no increase to teeth length due to increase of Vitamin C dose).

## Summary
We can conclude that:

- Supplement choice has no effect on teeth growth.
- Incrasing of the dose level will lead to increased teeth growth.

As long as the following assumptions are correct:

- That experiment was done with random assignment of quinea pigs to different dose levels and different suplement type.
- That sample members are representative of all population of guinea pigs.
- All the t-tests assumed the different variances for the 2 groups being compared.
