---
title: "Homework - Week 1"
author: "Martynas Asipauskas"
date: "Sunday, March 08, 2015"
output: html_document
---

This is a Homework solution / working document for Coursera - Statisical Inference course by John Hopkins University. [See assignment here](http://bcaffo.github.io/courses/06_StatisticalInference/homework/hw1.html#1)

# Problem set solutions
> Problem 1
> ([Source](http://bcaffo.github.io/courses/06_StatisticalInference/homework/hw1.html#4))

Overall probablity `p(A + B)` calculation is `p(A + B) = p(A) + p(B) - p(A intersect B)`
In this exercise unknown is: `p(A intersect B)`, by re-ordering equation above we get: `p(A intersect B) = p(A) + p(B) - p(A+B)`
Therefore the solution is: 0.04 or 4%.
```{r, echo=TRUE}
oneOfParents  <- 0.15
father        <- 0.1
mother        <- 0.09

(father + mother) - oneOfParents
```


> Problem 2
> ([Source](http://bcaffo.github.io/courses/06_StatisticalInference/homework/hw1.html#4))

As the overall shape is a `1 x 1` square, median percentile will be middle of it: `0.50`


> Problem 3
> ([Source](http://bcaffo.github.io/courses/06_StatisticalInference/homework/hw1.html#5))

The probability of tails is: `1 / (1 + d)`, whils the probablity of heads is: `d / (d + 1)`.
Therefore your total winnings will be: `Y * (1 / (1 + d))`, however at to calculate overall ammount won/lost, you have to deduct an ammount of your losses: `X * (d / (d + 1))`. Therefore total calculation will be: `Y * (1 / (1 + d)) - X * (d / (d + 1))`.


> Problem 4
> ([Source](http://bcaffo.github.io/courses/06_StatisticalInference/homework/hw1.html#6))

Variance is equal to: `Var(X)=E[X2]=(−4)^2∗.2+(1)^2∗.8`
```{r, echo=TRUE}
(−4)^2∗.2+(1)^2∗.8
```


> Problem 5
> ([Source](http://bcaffo.github.io/courses/06_StatisticalInference/homework/hw1.html#7))

Solution: `Var(Xˉ−Yˉ)=Var(Xˉ)+Var(Yˉ)=σ2/n+σ2/n`

> Problem 6
> ([Source](http://bcaffo.github.io/courses/06_StatisticalInference/homework/hw1.html#8))

Solution: `Var(X/σ)=Var(X)/σ2=1`

> Problem 7
> ([Source](http://bcaffo.github.io/courses/06_StatisticalInference/homework/hw1.html#9))

Solution: `Yes`

> Problem 8
> ([Source](http://bcaffo.github.io/courses/06_StatisticalInference/homework/hw1.html#10))

Variance formula is: `E[X2]−E[X]2`. Therefore solution is:
```{r, echo=TRUE}
p <- c(.1, .2, .3, .4)
x <- 2 : 5

sum(x ^ 2 * p) - sum(x * p) ^ 2
```
